\chapter{Conclusion}
Pokémon GO was a huge success compared to previous similar games, at its peak grossing for more than all other mobile games combined. We studied the possible reasons for its initial success and its subsequent downfall as Pokémon GO players suddenly went from being everywhere to being a rare sight. We also studied the potential effects playing the game had on its players' health, both physical and mental. The study was performed through a survey that garnered more than two thousand responses from players of all ages and activity levels.

Not surprisingly, the results from the survey indicate that the success of Pokémon GO largely comes from the combination of a Pokémon game on smart phones for availability and nostalgia, and a game that encouraged players to go outside and explore during summer when it was warm and they had the time. Players enjoyed following in the footsteps of the protagonists of the franchise, endeavoring to \emph{"catch 'em all"} and \emph{"become the very best"}, and many were excited about trying an augmented reality game, a genre they had not previously tried.

Unfortunately the game was lacking in content and things to do once players had caught all the monsters, and failed to keep most of its players engaged in the long term when they were unable to catch new monsters and school and work were back on their minds and schedules.

The game was very successful at bringing players out to play, when many of them would otherwise have stayed home participating in sedentary activities. Studies show that inactive lifestyles are the cause of many serious health issues, and increasing physical activity through intrinsically motivated activities is an effective way of preventing these kinds of lifestyle diseases. The game was particularly effective in making low activity players more active: where one in every two players were less physically active than the World Health Organization's recommended minimum weekly physical activity before Pokémon GO, only one in five were below the threshold while playing, while the group of players with a high level of activity more than tripled in size.

The increase in physical activity was not lasting, however, with most players falling back to their old ways after they stopped playing. Some players were inspired to start exercise by being exposed to exercise in a fun way, but for most players the increased physical activity was a side effect and not something to strive for without the motivation of Pokémon. Many players were also reckless while being consumed by the game, and risked playing while driving or while walking in traffic, or ventured into dangerous neighborhoods in pursuit of new adventures. Luckily, few serious accidents were caused by their actions.

While the game brought some disappointment and negative behavior, many sufferers of mental conditions such as depression or anxiety found the game helped them in dealing with their conditions through pushing them out of their safe spaces and into the world, where they were exposed to both exercise and human interaction. The strong community feeling of the game also helped foster many new friendships and improved existing relationships of all sorts.

The game showed promise, but ultimately failed to deliver in the long run. The style of game has the potential to have a serious impact on the collective health of society, as long as it manages to keep its players engaged for longer periods, and making a new success may be very possible if one can avoid the pitfalls that Pokémon GO fell into.


\chapter{Future Work}
This chapter presents some thoughts for possible future work on related subjects.

\section{Use of Pokémon GO in Serious Contexts}
Pokémon GO was a great success for players with low physical activity levels, and players suffering from mental conditions such as depression or anxiety. Further more focused and rigorous research on these effects could potentially uncover new findings and a more detailed picture of what exact aspects of the game is associated with the effects. If such studies are done, it is possible that the game or similar games can be used in treatment of patients with any of these conditions.

It is also possible that adding some special items or rewards to the game could motivate players to be more physically active, and what kind of items would be suitable for such a purpose is a possible subject for a study.

The game was also very successful at getting players to explore their local area, and could potentially be used to teach local history.

\section{Future of Pokémon GO and Similar Games}

While the AR mode in the game was not widely used, players enjoyed the pervasive aspects of the game, being able to navigate the map by moving in the real world and real world locations being tied to specific places in the game. The game's success was partly associated with the real-world integration and the technology used, and there should be potential for developing new, hugely successful games within these areas.

Entire communities such as the Pokémon GO Snap subreddit formed around the concept of taking fun, cute and interesting pictures using the in-game AR mode, and such pictures were also popular around the internet, so it would seem that AR technology still has much potential. The pervasive games genre is in growth, as seen in Chapter \ref{chapter:lit-study-similar-games}.
