\chapter{Follow-up Survey}
\label{appendix:follow-up}

The follow-up survey was distributed using the contact information left by players in the comments on the original survey.

All those contacted were asked the following questions:

\begin{enumerate}
\item Are you still playing?
\item If no to question 1, what made you quit and roughly when did this happen?
\item If no to question 1, are you likely to start playing again, and if so, what would make you come back?
\item Have you picked up any alternatives, and if so, how do you like them in comparison?
\item If yes to question 1, what would make you quit the game?
\item Of the time you reported playing, would you say the majority was active playing while physically active, or was more of the playing inactive (app open in pocket) or while being transported (e.g. via car or public transit)?
\item The questions on the last survey were almost exclusively worded from a positive perspective, but did you ever have any negative effects from playing, e.g. becoming less social or skipping workouts?
\item Have you ever bought coins in the game?
\item If yes, roughly how much have you spent?
\item If yes, what have you spent those coins on?
\item When third-party map services such as Pokévision were around, did you use them?
\item If you did use the map services, what was your goal (e.g. locating nests, catching specific rare mons, just seeing what's in the area, etc)?
\item If you used the maps, roughly for how long did you use them, and were you planning on stopping the use if/when the in-game tracker was fixed?
\item You originally reported that you increased your physical activity from 2 hours or less per week to 8 hours or less. Have you managed to keep this level of activity since?
\end{enumerate}

Additionally, players who were still playing when they responded to the follow-up survey were asked the following roughly one month after their response: \emph{"Are you still playing? If so, how much compared to last time?"}
